Survey of Cache Oblivious B-trees

### 2011-12-13 paper.tex
Edits
addressed Salman's comments on the co string b-tree section. moved co streaming b-tree section to be first.

The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
of hash data for the keys (known as the 
\emph{hashdata} layer), and a PMA of compressed values for the original key
 (known as the 
\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
the hashdata array. The centroid tree only contains a subset of the keys, so
each leaf points to the respective \emph{neighborhood} of the key being
searched for. After traversing the centroid tree and following the pointer into
the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
sequential scan is performed in the hashdata array layer to find $k$. Each
entry in the hashdata array contains the hash value of the corresponding key,
and a pointer to the compressed key value in the keydata layer. The keydata layer
is consulted to resolve hash collisions in the hashdata layer.

Consider the compacted\footnote{Compacted means all non-branching paths are
reduced to edges.} trie $T$ constructed over $D$ for string matching.  Consider
the vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants,
inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a
key $k$ in the compacted trie, you traverse the trie as follows: for each node
$y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on
into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
trie resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
obtained by rooting the tree at the centroid vertex $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

The centroid tree in the COSB-tree is layed out in memory using a modified vEB
(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. The
\emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
to rounding x down to the nearest power of a power of 2. This ensures trees
of equal size on all \emph{levels of detail}, as the centroid tree is traversed.

********* LEILANI: Can you provide a 5 word description of levels of details ***

********* LEILANI: What is a centroid vertex? ***

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for dynamic COSB-trees. Consider a sequence
of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using
standard front compression, let $\pi_{j}$ be the longest common prefix between
$k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:
\begin{equation}
  k_{1}||\pi_{2}||\sigma_{2}||\pi_{3}||\sigma_{3}||...||\pi_{j}||\sigma_{j}
\end{equation} 
where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
as well, making $k_{j}$ expensive to decode.

@@

>>>>

@@

The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
of hash data for the keys (known as the 
\emph{hashdata} layer), and a PMA of compressed values for the original key
 (known as the 
\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
the hashdata array. The centroid tree contains only hash values, and is designed such that
each leaf points to the respective \emph{neighborhood} of the key being
searched for. After traversing the centroid tree and following the pointer into
the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
sequential scan is performed in the hashdata array layer to find $k$. Each
entry in the hashdata array contains only the hash value of the corresponding key,
and a pointer to the compressed key value in the keydata layer. The keydata layer
is consulted to resolve hash collisions in the hashdata layer, and to obtain the original value of the key.

Consider the compacted\footnote{Compacted means all non-branching paths are
reduced to edges.} trie $T$ constructed over $D$ for string matching. There exists a
vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants,
inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a
key $k$ in the compacted trie, you traverse the trie as follows: for each node
$y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on
into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
trie resulting from excluding $y$ and its subtree, and call this the \emph{up trie} of $y$. The \emph{centroid tree} is
obtained by rooting the tree at the centroid vertex $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

The centroid tree in the COSB-tree is layed out in memory using a modified vEB
(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. Instead of using height $h/2$, the
\emph{hyperhyperfloors} of the weights of nodes are used as a guide for cutting the tree into
equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
to rounding x down to the nearest power of a power of 2.

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for dynamic COSB-trees. This approach 
increases the locality of the compressed data. Consider a sequence
of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using
standard front compression, let $\pi_{j}$ be the longest common prefix between
$k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:
\begin{equation}
  k_{1}||\pi_{2}||\sigma_{2}||\pi_{3}||\sigma_{3}||...||\pi_{j}||\sigma_{j}
\end{equation} 
where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
as well, making $k_{j}$ expensive to decode.
